---
title: "ComplexHeatmap Visualization with EasyProtein"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ComplexHeatmap Visualization with EasyProtein}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
suppressPackageStartupMessages({
  library(EasyProtein)
  library(tidyverse)
  library(SummarizedExperiment)
  library(ComplexHeatmap)
})
```

# ComplexHeatmap Visualization with EasyProtein

## 1. Tutorial metadata

- **Tutorial name**: ComplexHeatmap Visualization with EasyProtein
- **Tutorial type**: Heatmap-focused workflow
- **Target audience**: Users who already have EasyProtein-preprocessed proteomics data
- **Primary goal**: Show how to turn an EasyProtein `SummarizedExperiment` object into several common ComplexHeatmap visualizations
- **Expected outcome**: You will learn how to:
  1. load data with `rawdata2se()`
  2. build sample-level and condition-level heatmap matrices
  3. select protein subsets from PCA or user-defined gene lists
  4. add column and row annotations
  5. split rows by unsupervised clusters
  6. use `plot_heatmap_withline()` for cluster-level trend summaries
  7. export heatmaps with `saveplot()`

## 2. Load example data and build `SummarizedExperiment`

This tutorial uses the example file shipped with EasyProtein. In a real
analysis, replace `exp_file` with your own EasyProtein-compatible protein
matrix, then keep the downstream code unchanged.

```{r load-data}
exp_file <- system.file(
  "extdata",
  "mouse_multi-organ_DIA_report.pg_matrix.tsv.gz",
  package = "EasyProtein"
)

stopifnot(file.exists(exp_file))

se_obj <- rawdata2se(exp_file)
se <- se_obj$se
se
```

## 3. Add sample metadata

ComplexHeatmap becomes most useful when the columns can be split or annotated
by biological variables. Here we derive `tissue`, `age`, and `sex` from the
example `condition` column, following the same data preparation used in the
quick start and multi-sample tutorials.

```{r add-metadata}
se$tissue <- stringr::str_extract(se$condition, "[A-Za-z]+")
se$age <- stringr::str_extract(se$condition, "\\dw")
se$sex <- stringr::str_extract(se$condition, "\\w$")

sample_table <- as.data.frame(SummarizedExperiment::colData(se)) %>%
  dplyr::select(sample, condition, rep, tissue, age, sex)

head(sample_table, 10)
```

For compact figures, we focus on one coherent subset: male heart samples across
three ages. This keeps the biological interpretation simple while preserving
replicate-level structure.

```{r subset-data}
se_focus <- se[, se$tissue == "heart" & se$sex == "M"]
se_focus$age <- factor(se_focus$age, levels = c("1w", "4w", "8w"))
se_focus$condition <- factor(se_focus$condition, levels = unique(se_focus$condition))

rep_levels <- unique(as.character(se_focus$rep))
rep_numeric <- suppressWarnings(as.numeric(rep_levels))
if (all(!is.na(rep_numeric))) {
  rep_levels <- rep_levels[order(rep_numeric)]
} else {
  rep_levels <- sort(rep_levels)
}
se_focus$rep <- factor(se_focus$rep, levels = rep_levels)

as.data.frame(SummarizedExperiment::colData(se_focus)) %>%
  dplyr::select(sample, condition, rep, tissue, age, sex)
```

## 4. Build heatmap matrices from EasyProtein objects

EasyProtein stores several assays in the `SummarizedExperiment` object. For
heatmaps, two views are especially common:

- **sample-level matrix**: one column per biological replicate
- **condition-level matrix**: one column per condition, usually the mean or median across replicates

For visual stability, the examples below select the top variable proteins
instead of plotting every detected protein in the vignette.

```{r prepare-matrices}
z_df <- se2scale(se_focus, rescale = TRUE)

z_mtx <- as.matrix(z_df[, colnames(se_focus), drop = FALSE])
rownames(z_mtx) <- rownames(se_focus)
z_mtx[!is.finite(z_mtx)] <- 0

top_n <- min(300, nrow(z_mtx))
row_var <- matrixStats::rowVars(z_mtx)
names(row_var) <- rownames(z_mtx)
top_variable_genes <- names(sort(row_var, decreasing = TRUE))[seq_len(top_n)]

z_top <- z_mtx[top_variable_genes, , drop = FALSE]

condition_mean_mtx <- calc_gene_mean_by_condition(
  se = se_focus,
  assay_name = "conc",
  method = "mean",
  condition_col = "age"
)

condition_mean_z <- scale_mtx(condition_mean_mtx)
condition_mean_z[!is.finite(condition_mean_z)] <- 0
condition_mean_top <- condition_mean_z[top_variable_genes, , drop = FALSE]

dim(z_top)
dim(condition_mean_top)
```

If your input data contain duplicated gene symbols, keep row names as unique
protein IDs for computation. You can still display gene symbols later by adding
row annotations or by selecting representative rows after aggregation.

## 5. Sample-level heatmap

The first heatmap keeps biological replicates as separate columns. This view is
useful for checking whether replicates behave consistently inside each age group.

```{r sample-heatmap, fig.width=7, fig.height=6}
ht_sample <- ComplexHeatmap::Heatmap(
  z_top,
  name = "Z-score",
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  show_row_names = FALSE,
  show_column_names = FALSE,
  column_split = se_focus$age,
  column_title = NULL,
  use_raster = TRUE
)

ht_sample
```

## 6. Condition-level heatmap

When the question is about the overall pattern across conditions, average the
replicates first. This makes the heatmap smaller and easier to interpret.

```{r condition-heatmap, fig.width=5, fig.height=6}
ht_condition <- ComplexHeatmap::Heatmap(
  condition_mean_top,
  name = "Z-score",
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  show_row_names = FALSE,
  show_column_names = TRUE,
  column_title = NULL,
  use_raster = TRUE
)

ht_condition
```

Use the sample-level heatmap to inspect replicate behavior. Use the
condition-level heatmap to communicate the main biological trend.

## 7. Add column annotations

Column annotations make the experimental design explicit. The example below
colors age groups and replicate numbers above the sample-level heatmap.

```{r column-annotation, fig.width=7, fig.height=6}
age_col <- c(
  "1w" = "#1F77B4",
  "4w" = "#2CA02C",
  "8w" = "#D62728"
)

rep_levels <- levels(se_focus$rep)
rep_col <- setNames(
  grDevices::hcl.colors(length(rep_levels), palette = "Dark 3"),
  rep_levels
)

top_anno <- ComplexHeatmap::HeatmapAnnotation(
  age = se_focus$age,
  rep = se_focus$rep,
  col = list(
    age = age_col,
    rep = rep_col
  ),
  show_annotation_name = TRUE
)

ht_anno <- ComplexHeatmap::Heatmap(
  z_top,
  name = "Z-score",
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  show_row_names = FALSE,
  show_column_names = FALSE,
  top_annotation = top_anno,
  column_split = se_focus$age,
  column_title = NULL,
  use_raster = TRUE
)

ht_anno
```

## 8. Heatmap for selected proteins

You often do not want an unsupervised heatmap of all proteins. Instead, select
proteins from a biological hypothesis, PCA contributors, DEG results, or any
custom gene list.

### 8.1 Select proteins from PCA contributors

This example identifies proteins that contribute strongly to PC1 and draws a
focused heatmap. Positive and negative contributors are used as row splits.

```{r pca-contributors}
conc_mtx <- SummarizedExperiment::assay(se_focus, "conc")
pca_res <- FactoMineR::PCA(t(conc_mtx), scale.unit = TRUE, graph = FALSE)

pc1_genes <- dplyr::bind_rows(
  get_pc_contributors(pca_res, se_focus, pc = 1, n = 30)$positive,
  get_pc_contributors(pca_res, se_focus, pc = 1, n = 30)$negative
) %>%
  dplyr::filter(gene %in% rownames(z_mtx)) %>%
  dplyr::distinct(gene, .keep_all = TRUE)

pc1_label_genes <- pc1_genes %>%
  dplyr::group_by(type) %>%
  dplyr::slice_head(n = 5) %>%
  dplyr::ungroup() %>%
  dplyr::pull(gene)

head(pc1_genes)
```

```{r pca-heatmap, fig.width=7, fig.height=5}
pc1_mtx <- z_mtx[pc1_genes$gene, , drop = FALSE]
pc1_row_labels <- ifelse(rownames(pc1_mtx) %in% pc1_label_genes, rownames(pc1_mtx), "")

ht_pc1 <- ComplexHeatmap::Heatmap(
  pc1_mtx,
  name = "Z-score",
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  row_split = pc1_genes$type,
  show_row_names = TRUE,
  row_labels = pc1_row_labels,
  show_column_names = FALSE,
  top_annotation = top_anno,
  column_split = se_focus$age,
  column_title = NULL,
  use_raster = TRUE
)

ht_pc1
```

### 8.2 Select proteins from a custom gene list

For user-defined markers, keep only genes found in the matrix. The same pattern
works for lists imported from CSV, Excel, enrichment results, or manual curation.

```{r custom-gene-list, eval=FALSE}
target_genes <- c("Actb", "Gapdh", "Myh6")
target_genes <- intersect(target_genes, rownames(z_mtx))

ht_targets <- ComplexHeatmap::Heatmap(
  z_mtx[target_genes, , drop = FALSE],
  name = "Z-score",
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  show_row_names = TRUE,
  show_column_names = FALSE,
  top_annotation = top_anno,
  column_split = se_focus$age,
  column_title = NULL
)

ht_targets
```

## 9. Unsupervised row clustering

For exploratory analysis, cluster the condition-level matrix first, then use the
cluster assignment to split rows in the replicate-level heatmap.

```{r kmeans-clustering}
set.seed(1)

k <- 4
km <- stats::kmeans(condition_mean_top, centers = k, nstart = 20)

row_cluster_df <- tibble::tibble(
  gene = rownames(condition_mean_top),
  km_cluster = paste0("km", km$cluster)
) %>%
  dplyr::arrange(km_cluster)

head(row_cluster_df)
```

```{r cluster-heatmap, fig.width=7, fig.height=6}
cluster_genes <- row_cluster_df$gene

ht_cluster <- ComplexHeatmap::Heatmap(
  z_mtx[cluster_genes, , drop = FALSE],
  name = "Z-score",
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  row_split = row_cluster_df$km_cluster,
  show_row_names = FALSE,
  show_column_names = FALSE,
  top_annotation = top_anno,
  column_split = se_focus$age,
  column_title = NULL,
  use_raster = TRUE
)

ht_cluster
```

The number of clusters should be chosen according to the biological question.
Start with a small value such as 4 to 6, then inspect whether the resulting
groups are interpretable.

### 9.1 Enrichment analysis for each k-means cluster

After the heatmap clusters are defined, each cluster can be treated as a gene
set for functional interpretation. The code below runs GO enrichment for every
`km_cluster` and prints the top terms per cluster when `clusterProfiler` and the
mouse annotation database are available.

```{r km-enrichment}
cluster_gene_list <- split(row_cluster_df$gene, row_cluster_df$km_cluster)

if (
  requireNamespace("clusterProfiler", quietly = TRUE) &&
  requireNamespace("org.Mm.eg.db", quietly = TRUE)
) {
  km_go_list <- purrr::imap(cluster_gene_list, function(genes, cluster_name) {
    genes <- unique(stats::na.omit(genes))

    if (length(genes) < 5) {
      return(NULL)
    }

    res <- tryCatch(
      enrichment_analysis(
        gene = genes,
        db = "GO",
        species = "mouse",
        keyType = "SYMBOL",
        qvalue_cutoff = 0.05
      ),
      error = function(e) {
        message(cluster_name, ": enrichment skipped - ", conditionMessage(e))
        NULL
      }
    )

    if (is.null(res) || nrow(res) == 0) {
      return(NULL)
    }

    dplyr::mutate(as.data.frame(res), km_cluster = cluster_name, .before = 1)
  })

  km_go_df <- dplyr::bind_rows(km_go_list)

  if (nrow(km_go_df) > 0) {
    km_go_df %>%
      dplyr::group_by(km_cluster) %>%
      dplyr::slice_head(n = 5) %>%
      dplyr::ungroup()
  } else {
    message("No enriched GO terms were found for the current k-means clusters.")
  }
} else {
  message("Install 'clusterProfiler' and 'org.Mm.eg.db' to run per-cluster GO enrichment.")
}
```

## 10. Add row annotations

Row annotations are useful when you have feature-level labels such as protein
class, DEG direction, pathway membership, or cluster identity.

```{r row-annotation, fig.width=8, fig.height=6}
row_cluster_df <- row_cluster_df %>%
  dplyr::mutate(direction = ifelse(km_cluster %in% c("km1", "km2"), "early_high", "late_high"))

direction_levels <- unique(row_cluster_df$direction)
direction_col <- setNames(
  grDevices::hcl.colors(length(direction_levels), palette = "Set 2"),
  direction_levels
)

right_anno <- ComplexHeatmap::rowAnnotation(
  direction = row_cluster_df$direction,
  col = list(direction = direction_col),
  show_annotation_name = TRUE
)

ht_row_anno <- ComplexHeatmap::Heatmap(
  z_mtx[row_cluster_df$gene, , drop = FALSE],
  name = "Z-score",
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  row_split = row_cluster_df$km_cluster,
  show_row_names = FALSE,
  show_column_names = FALSE,
  top_annotation = top_anno,
  right_annotation = right_anno,
  column_split = se_focus$age,
  column_title = NULL,
  use_raster = TRUE
)

ht_row_anno
```

In a real project, replace the toy `direction` labels with actual annotations,
for example `DEGs_types`, pathway class, subcellular location, or cluster labels
from `se2gene_group()`.

## 11. Cluster heatmap with trend summary lines

EasyProtein includes `plot_heatmap_withline()`, a ComplexHeatmap wrapper that
adds a right-side line panel for each row cluster. This is useful when you want
to show both the per-sample heatmap and the average trend of each cluster.

```{r heatmap-withline, fig.width=9, fig.height=6}
ht_line <- plot_heatmap_withline(
  mat = z_mtx[row_cluster_df$gene, colnames(se_focus), drop = FALSE],
  row_split = row_cluster_df$km_cluster,
  column_split = se_focus$age,
  top_annotation = top_anno,
  show_column_names = FALSE,
  set.md = "mean"
)

ht_line
```

This plot is best used after you have a meaningful row grouping. The grouping
can come from k-means, Mfuzz, time-course clustering, DEG direction, pathway
sets, or any manually curated protein class.

## 12. Export heatmaps

`saveplot()` can save ComplexHeatmap objects to the standard EasyProtein figure
folders. The code below is disabled during vignette rendering so that package
builds do not write output files.

```{r save-heatmaps, eval=FALSE}
saveplot(
  object = ht_sample,
  filenames = "sample_level_heatmap",
  width = 4,
  height = 4
)

saveplot(
  object = ht_cluster,
  filenames = "clustered_heatmap",
  width = 5,
  height = 5
)

saveplot(
  object = ht_line,
  filenames = "clustered_heatmap_with_trend_lines",
  width = 6,
  height = 4
)
```

For publication figures, adjust `width`, `height`, row names, annotation colors,
and cluster numbers after inspecting the first exported draft.

## 13. Practical templates

### 13.1 Use your own preprocessed EasyProtein object

```{r own-data-template, eval=FALSE}
se <- readRDS("Rds/my_easyprotein_se.rds")

se$condition <- factor(se$condition, levels = unique(se$condition))

z_df <- se2scale(se, rescale = TRUE)
z_mtx <- as.matrix(z_df[, colnames(se), drop = FALSE])
rownames(z_mtx) <- rownames(se)
z_mtx[!is.finite(z_mtx)] <- 0

row_var <- matrixStats::rowVars(z_mtx)
names(row_var) <- rownames(z_mtx)
top_n <- min(500, nrow(z_mtx))
top_genes <- names(sort(row_var, decreasing = TRUE))[seq_len(top_n)]

ComplexHeatmap::Heatmap(
  z_mtx[top_genes, , drop = FALSE],
  name = "Z-score",
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  column_split = se$condition,
  show_row_names = FALSE,
  show_column_names = FALSE,
  use_raster = TRUE
)
```

### 13.2 Aggregate duplicated gene symbols before plotting

Some proteomics reports contain multiple protein IDs for the same gene symbol.
When you want a gene-level heatmap, aggregate the matrix before plotting.

```{r aggregate-template, eval=FALSE}
scale_df <- se2scale(se, rescale = TRUE) %>%
  dplyr::mutate(gene_symbol = stringr::str_extract(Genes, "^[^;]+")) %>%
  dplyr::filter(!is.na(gene_symbol), gene_symbol != "") %>%
  dplyr::group_by(gene_symbol) %>%
  dplyr::summarise(
    dplyr::across(dplyr::all_of(colnames(se)), mean, na.rm = TRUE),
    .groups = "drop"
  )

gene_z_mtx <- as.matrix(scale_df[, colnames(se), drop = FALSE])
rownames(gene_z_mtx) <- scale_df$gene_symbol
```

### 13.3 Plot DEG proteins only

```{r deg-template, fig.width=7, fig.height=5}
deg_res <- se2DEGs(
  se = se_focus,
  compare_col = "age",
  ref = "1w",
  cmp = "8w",
  logFC_cutoff = 1,
  adj_p_cutoff = 0.05
)

deg_genes <- deg_res %>%
  dplyr::filter(DEGs_types %in% c("UP", "DOWN")) %>%
  dplyr::pull(gene) %>%
  unique()

deg_genes <- intersect(deg_genes, rownames(z_mtx))

if (length(deg_genes) == 0) {
  deg_genes <- deg_res %>%
    dplyr::arrange(adj.P.Val, dplyr::desc(abs(logFC))) %>%
    dplyr::slice_head(n = 40) %>%
    dplyr::pull(gene) %>%
    intersect(rownames(z_mtx))
}

ht_deg <- ComplexHeatmap::Heatmap(
  z_mtx[deg_genes, , drop = FALSE],
  name = "Z-score",
  row_split = deg_res$DEGs_types[match(deg_genes, deg_res$gene)],
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  column_split = se_focus$age,
  top_annotation = top_anno,
  show_row_names = FALSE,
  show_column_names = FALSE
)

ht_deg
```

## 14. Recommended heatmap workflow

1. Build `se` with `rawdata2se()` or load a saved EasyProtein `SummarizedExperiment`.
2. Add explicit metadata columns to `colData(se)`.
3. Use `se2scale()` for sample-level z-score heatmaps.
4. Use `calc_gene_mean_by_condition()` plus `scale_mtx()` for condition-level heatmaps.
5. Select a biologically meaningful feature set before fine-tuning aesthetics.
6. Add `HeatmapAnnotation()` and `rowAnnotation()` only after the matrix and ordering are correct.
7. Use `plot_heatmap_withline()` when row clusters need trend summaries.
8. Export final figures with `saveplot()`.
