---
title: "Hybrid Temporal Signature Clustering"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Hybrid Temporal Signature Clustering}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

Hybrid Temporal Signature Clustering (HTSC) identifies coordinated temporal
protein response programs across one or multiple biological time series. It is
designed for experiments such as multiple cell lines measured across the same
ordered time points.

For protein \(g\), series \(s\), time point \(t\), and replicate \(r\), let
\(Y_{g,s,t,r}\) be the normalized abundance. Values are transformed as

\[
X_{g,s,t,r}=\log_2(Y_{g,s,t,r}+\epsilon),
\]

where \(\epsilon\) is half of the minimum positive abundance unless specified
by the user.

Within each series, HTSC compares every post-baseline time point with the
baseline:

\[
L_{g,s,t}=\widehat{\beta}_{g,s,t}-\widehat{\beta}_{g,s,0}.
\]

The contrast is tested with limma and adjusted P values are denoted
\(q_{g,s,t}\). The clustering matrix is a significance-gated log2FC matrix:

\[
Z_{g,s,t} =
\begin{cases}
L_{g,s,t}, & q_{g,s,t}\leq \alpha \ \land\ |L_{g,s,t}|\geq \delta,\\
0, & \mathrm{otherwise}.
\end{cases}
\]

The baseline is fixed to \(Z_{g,s,0}=0\). A protein is treated as dynamic when

\[
\sum_s \sum_{t>0} \mathbf{1}(Z_{g,s,t}\neq 0)\geq m.
\]

The default is \(m=1\), so a protein enters dynamic clustering if it changes
significantly at least once in any series. Proteins that do not pass this rule
are reported as `Flat_or_not_significant`.

The multi-series temporal signature is

\[
\mathbf{z}_g =
\left[
Z_{g,1,0},\ldots,Z_{g,1,T_1},
\ldots,
Z_{g,S,0},\ldots,Z_{g,S,T_S}
\right]^\top .
\]

HTSC clusters dynamic proteins using a hybrid distance:

\[
D(g,h)=
w_{\mathrm{shape}}D_{\mathrm{shape}}^{*}(g,h)
+w_{\mathrm{amp}}D_{\mathrm{amp}}^{*}(g,h)
+w_{\mathrm{trend}}D_{\mathrm{trend}}^{*}(g,h)
+w_{\mathrm{time}}D_{\mathrm{time}}^{*}(g,h).
\]

The shape term compares standardized profiles,

\[
D_{\mathrm{shape}}(g,h)=
\frac{1-\mathrm{cor}(\tilde{\mathbf{z}}_g,\tilde{\mathbf{z}}_h)}{2}.
\]

The amplitude term preserves response strength,

\[
D_{\mathrm{amp}}(g,h)=\|\mathbf{z}_g-\mathbf{z}_h\|_2.
\]

The trend term compares first differences within each series,

\[
\Delta_{g,s,t}=Z_{g,s,t}-Z_{g,s,t-1},
\qquad
D_{\mathrm{trend}}(g,h)=
\|\boldsymbol{\Delta}_g-\boldsymbol{\Delta}_h\|_2.
\]

The timing term compares significant-response positions:

\[
D_{\mathrm{time}}(g,h)=
\frac{
\sum_p \mathbf{1}(b_{g,p}\neq b_{h,p})
}{
\sum_p \mathbf{1}(b_{g,p}\lor b_{h,p})
}.
\]

Each distance component is divided by its empirical 95th percentile and capped
at 1:

\[
D_k^{*}(g,h)=
\min\left(\frac{D_k(g,h)}{Q_{0.95}(D_k)},1\right).
\]

The default weights are

\[
(w_{\mathrm{shape}},w_{\mathrm{amp}},w_{\mathrm{trend}},w_{\mathrm{time}})
=(0.35,0.25,0.25,0.15).
\]

Dynamic proteins are clustered with Ward hierarchical clustering. When
`k = "auto"`, the number of clusters is selected by maximizing average
silhouette width:

\[
K^*=\arg\max_{K\in\mathcal{K}}
\frac{1}{|\mathcal{G}_{dyn}|}
\sum_{g\in\mathcal{G}_{dyn}}\mathrm{sil}_K(g).
\]

Example usage:

```{r, eval = FALSE}
library(EasyProtein)

res <- temporalSignatureClustering(
  se,
  assay_name = "conc",
  time_col = "time",
  series_col = "cell_line",
  baseline = 0,
  k = "auto",
  k_range = 4:10,
  p_adj_cutoff = 0.05,
  logFC_cutoff = 0.5,
  min_significant_timepoints = 1
)

head(res$gene_cluster)
res$cluster_summary
res$cluster_change
```

The returned object includes gene-level clusters, all-gene assignments with a
background group, limma time-point tests, centroid trajectories, and gene- and
cluster-level significant response times.
