---
title: "Quick Start with EasyProtein"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start with EasyProtein}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(EasyProtein)
library(tidyverse)
library(SummarizedExperiment)
```

# Quick Start with EasyProtein

## 1. Tutorial metadata

- **Tutorial name**: Quick Start with EasyProtein  
- **Tutorial type**: Minimal end-to-end workflow  
- **Target audience**: First-time users of EasyProtein  
- **Primary goal**: Show how to complete a basic proteomics analysis workflow with minimal input and minimal parameter tuning  
- **Expected outcome**: You will learn how to:
  1. load example data
  2. create a standard EasyProtein analysis object
  3. inspect data quality
  4. visualize sample structure
  5. perform a simple two-group comparison (**Liver, 1w vs 8w, male only**)
  6. interpret core outputs

## 2. Load example data and build `SummarizedExperiment`

This tutorial uses the file `inst/extdata/mouse_multi-organ_DIA_report.pg_matrix.tsv.gz`.

```{r load-data}
exp_file <- system.file(
  "extdata",
  "mouse_multi-organ_DIA_report.pg_matrix.tsv.gz",
  package = "EasyProtein"
)

stopifnot(file.exists(exp_file))

se_obj <- rawdata2se(exp_file)
se <- se_obj$se
se
```

## 3. Add sample metadata

`rawdata2se()` already creates basic `condition` and `rep`. Here we derive `tissue`, `age`, and `sex` from `condition` as requested.

```{r add-metadata}
se$tissue <- str_extract(se$condition,'[A-Za-z]+')
se$age <- str_extract(se$condition,'\\dw')
se$sex <- str_extract(se$condition,'\\w$')

as.data.frame(SummarizedExperiment::colData(se)) %>%
  dplyr::select(sample, condition, rep, tissue, age, sex) %>%
  head()
```

## 4. Quick QC

We inspect three basic QC aspects:

1. sample intensity distribution
2. missing values per sample
3. detected protein number per sample

```{r qc-plots, fig.width=10, fig.height=8}
keep <- se$tissue == "live" &
  se$sex == "M" &
  se$age %in% c("1w", "8w")

se_sub <- se[, keep]


p_density <- plotSE_density(se_sub)
p_missing <- plotSE_missing_value(se_sub)
p_detected <- plotSE_protein_number(se_sub)

p_density
p_missing
p_detected
```

## 5. Visualize sample structure with PCA

Use concentration matrix (`assay = "conc"`) and color by tissue.

```{r pca, fig.width=7, fig.height=5}
conc_mtx <- SummarizedExperiment::assay(se_sub, "conc")

pca.res <- FactoMineR::PCA(t(conc_mtx), graph = FALSE)
pca.df <- as.data.frame(pca.res$ind$coord)
pca.df$sample <- rownames(pca.df)

meta_df <- as.data.frame(SummarizedExperiment::colData(se_sub)) 

pca.df <- dplyr::left_join(pca.df, meta_df, by = "sample")

plot_pca(
  pca.df = pca.df,
  pca.res = pca.res,
  colorby = "age",
  label = "sample"
)
```

## 6. Two-group comparison: Liver, 1w vs 8w, male only

We subset samples to:

- `tissue == "Liver"`
- `sex == "m"`
- `age %in% c("1w", "8w")`

Then run unpaired limma-based DEG analysis via `se2DEGs()`.

```{r deg-run}
deg_res <- se2DEGs(
  se = se_sub,
  compare_col = "age",
  ref = "1w",
  cmp = "8w",
  logFC_cutoff = 1,
  adj_p_cutoff = 0.05
)

head(deg_res)
table(deg_res$DEGs_types)
```

## 7. Interpret core outputs

The DEG result table contains:

- **effect size**: `logFC` (positive means higher in `8w` vs `1w`)
- **statistical significance**: `P.Value`, `adj.P.Val`
- **classification**: `DEGs_types` (`UP`, `DOWN`, `NS`)
- **group summaries**: `median_1w`, `median_8w`

Minimal interpretation workflow:

1. prioritize proteins with `adj.P.Val <= 0.05`
2. use `|logFC| >= 1` to focus on stronger changes
3. read `median_1w` and `median_8w` together with `logFC`

```{r top-hits}
top_hits <- deg_res %>%
  dplyr::arrange(adj.P.Val, dplyr::desc(abs(logFC))) %>%
  dplyr::select(gene, logFC, adj.P.Val, DEGs_types, starts_with("median_"))

head(top_hits, 20)
```

## 8. Save results (optional)

```{r save-results, eval=FALSE}
write.csv(deg_res, "Liver_male_1w_vs_8w_DEGs.csv", row.names = FALSE)
```
